Deep Learning for Dynamic NFT Valuation

### 2023-12-08 paper/main.tex
update paper title
write full paper

% to typeset URLs, URIs, and DOIs
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{url}
\usepackage{graphicx}
\usepackage[english]{babel}

@@

\title{Machine Learning for NFT Price Predictions}

\author{Mingxuan He\thanks{mingxuanh@uchicago.edu, Chicago, IL, United States of America}}
\date{\today}

@@

\begin{abstract}
In this proposal, I outline the motivation, data, methods, and preliminary results of my project. The goal of this project is to predict the price of NFTs using classical machine learning and deep learning methods. Training and testing data include transactions data from the Ethereum blockchain via Dune Analytics, and NFT collection data from OpenSea, the largest NFT marketplace. The results will be presented in a paper and a presentation. The code will be written in Python utilizing packages including scikit-learn and tensorflow.\\

\textbf{Keywords:} NFTs, machine learning, onchain data, predictive modeling

@@

In this paper, I apply machine learning methods to predict the price of picture NFTs. Specifically, I train models on market prices and rarity data obtained from the Ethereum blockchain. The goal is to estalish a model framework that provides accurate estimates of NFT prices. Such a model has various applications, including but not limited to onchain trading and bidding contracts, risk management for crypto portfolios, and fair valuation of staked/collateralized NFTs for decentralized finance (DeFi) applications. In addition, the model can be used to price large batches of NFTs quickly, which is useful for AI-generated NFTs.

@@

\section{Literature Review}
\label{sec: lit_review}
% market analysis
The fast-growing market of NFTs has attracted interest from various academic disciplines. In the economics and finance literature, various authors have discussed the price dynamics of the NFT market. A comprehensive work by \citet{nadini2021mapping} mapped out important statistical features of NFT markets, revealing that mean prices, sales per asset, sales per collection all follow power-law distributions. \citet{ante2022non} found that NFT sales are triggered by price shocks in Bitcoin (BTC) and active NFT wallets are reduced by price shocks in  Ether (ETH). Similarly, \citet{dowling2022non} found co-movement between cryptocurrencies and NFTs through wavelet coherence analysis. On the buyer side, \citet{kong2021alternative} that well-connected and experienced investors generally pay lower prices for NFTs. These results support the general belief that the NFT market is not fully efficient, and there is room for improvement in pricing NFTs.

@@

% ML
In the applied machine learning literature, there have been notable attempts to apply regression models and neural networks to predict the price of NFTs. \citet{nadini2021mapping} used a linear regression model with network features such as the buyer's and seller's degree and PageRank centrality, principal components of the NFT's visual features, as well as market data such as the past median price of primary and secondary sales.

The literature also identified other features such as search trends \citep{jain2022nft,kaneko2021time}, social media influence e.g. Twitter \citep{kapoor2022tweetboost}.

@@

@@

\subsection{Exploratory analysis}
% TODO: summary stats with sample size
\begin{figure}[H]
    \includegraphics[width=\textwidth]{../figures/price_date.png}
    \caption{Trade Price by Date}
\end{figure}

\begin{figure}[H]
    \includegraphics[width=\textwidth]{../figures/price_rarity.png}
    \caption{Trade Price by Rarity Rank}
\end{figure}

\section{Methodology}
\label{sec: method}
For benchmarking, I will use an OLS model. For the main model, I will use a neural network with a fully connected layer. I will also explore other deep learning methods, including convolutional neural networks (CNNs) and recurrent neural networks (RNNs). This is currently work in progress.

\section{Results}
\label{sec: results}

@@

@@

>>>>

% to typeset URLs, URIs, and DOIs
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{url}
\usepackage{graphicx}
%\usepackage[nolists]{endfloat}
\usepackage[english]{babel}

@@

\title{Deep Learning for Dynamic NFT Valuation}

\author{Mingxuan He
    \thanks{mingxuanh@uchicago.edu, Chicago, IL, United States of America}}
\date{\today}

@@

\begin{abstract}
    I study the price dynamics of non-fungible tokens (NFTs) and propose a deep learning framework for dynamic valuation of NFTs. I use data from the Ethereum blockchain and OpenSea to train a deep learning model on historical trades, market trends, and traits/rarity features of Bored Ape Yacht Club NFTs. After hyperparameter tuning, the model is able to predict the price of NFTs with high accuracy. I propose an application framework for this model using zero-knowledge machine learning (zkML) and discuss its potential use cases in the context of decentralized finance (DeFi) applications.\\

    \textbf{Keywords:} NFTs, deep learning, onchain data, zkML

@@

In this paper, I apply machine learning methods to value the price of picture NFTs. Specifically, I train models on market prices and rarity data obtained from the Ethereum blockchain. The goal is to estalish a model framework that provides accurate estimates of NFT prices. Such a model has various applications, including but not limited to onchain trading and bidding contracts, risk management for crypto portfolios, and fair valuation of staked/collateralized NFTs for decentralized finance (DeFi) applications. In addition, the model can be used to price large batches of NFTs quickly, which is useful for AI-generated NFTs.

@@

\section{Literature Review}
\label{sec: lit_review}
% market analysis
The fast-growing market of NFTs has attracted interest from various academic disciplines. In the economics and finance literature, various authors have discussed the price dynamics of the NFT market. A comprehensive work by \citet{nadini2021mapping} mapped out important statistical features of NFT markets, revealing that mean prices, sales per asset, sales per collection all follow power-law distributions. \citet{ante2022non} found that NFT sales are triggered by price shocks in Bitcoin (BTC) and active NFT wallets are reduced by price shocks in  Ether (ETH). Similarly, \citet{dowling2022non} found co-movement between cryptocurrencies and NFTs through wavelet coherence analysis. On the buyer side, \citet{kong2021alternative,oh2022investor} found that well-connected and experienced investors generally earn higher returns on NFT trades. These results support the general belief that the NFT market is not fully efficient, and there is room for improvement in pricing NFTs.

@@

% ML
In the applied machine learning literature, there have been notable attempts to train predictive models on visual features of NFTs. \citet{nadini2021mapping} built a linear regression model with extracted principal components of the NFT's image, features of the trading network, and market history. \citet{seyhan2023nft} and \citet{costa2023show} used multimodal data including NFT images and text characteristics while excluding financial features in an effort to predict both primary and secondary sale prices. The literature also identified other features such as search trends \citep{jain2022nft,kaneko2021time}, social media influence e.g. Twitter \citep{kapoor2022tweetboost}. My proposed framework contributes to the literature by incorporating a large set of tabular features and build a deep learning model tailored to each collection. This approach is more scalable and can be applied to a wide range of collections for secondary sale price prediction and valuation, especially in fully decentralized applications where image data is costly to store and transfer onchain.

@@

In this paper, I use the Bored Ape Yacht Club (BAYC) collection as an example. The Bored Ape Yacht Club is a well-known collection of 10,000 unique Bored Ape NFTs. The collection was created by Yuga Labs in April 2021. BAYC has a total trading volume of 1.4 million ETH (\$2.2 billion USD) as of December 2023. BAYC is favored by many celebrities and has a large community of collectors.

@@

\subsection{Data description}
As of September 30, 2023, there has been 43,383 BAYC trades recorded in the Dune \texttt{nft.trades} database. It is worth noting that trades conducted outside of marketplaces, such as through private over-the-counter agreements, are not recorded, however those agreements occur not as frequently. Here I outline the features and target variables used in this project. The features are divided into three categories: market, traits \& rarity, and last trade. The features are described in Table \ref{tab: feature_columns}.
\begin{table}[H]
    \centering
    \caption{Feature Columns Description (Bored Ape Yacht Club)}
    \begin{tabular}{|l|l|l|}
    \hline
    \textbf{Feature Category} & \textbf{Feature} & \textbf{Description} \\ \hline

    \multirow{4}{*}{Market} & volume\_eth\_lag1 & daily market volume (ETH) \\ \cline{2-3} 
    & price\_p5\_eth\_lag1 & daily 5-percentile price (ETH) \\ \cline{2-3} 
    & price\_max\_eth\_lag1 & daily highest price (ETH) \\ \cline{2-3} 
    & price\_min\_eth\_lag1 & daily minimum (floor) price (ETH)\\ \hline

    \multirow{8}{*}{Traits \& Rarity} & rarity\_rank & rarity rank measured by OpenRarity \\ \cline{2-3} 
    & Background\_count & number of items with the same \textit{Background} trait \\ \cline{2-3} 
    & Mouth\_count & number of items with the same \textit{Mouth} trait \\ \cline{2-3} 
    & Eyes\_count & number of items with the same \textit{Eyes} trait\\ \cline{2-3} 
    & Fur\_count & number of items with the same \textit{Fur} trait \\ \cline{2-3}
    & Clothes\_count & number of items with the same \textit{Clothes} trait \\ \cline{2-3} 
    & Earring\_count & number of items with the same \textit{Earring} trait \\ \hline

    \multirow{2}{*}{Last Trade} & last\_trade\_timediff & time since the last time this item was traded \\ \cline{2-3} 
    & last\_trade\_price & price at the last time this item was traded\\ \hline
    \end{tabular}

    \label{tab: feature_columns}
\end{table}

Note that all market data (volume and prices) are at daily frequency and lagged by 1 day. The traits and rarity data are static and do not change over time. The last trade data is dynamic and changes with each trade. For each trait, I use the count of items in the collection which shares the same trait (e.g. \texttt{Fur="tan"} corresponds to \texttt{Fur\_count=626}), with lower counts signaling higher rarity of the specific trait. The outcome variable is the trade price of the NFT (in ETH).

\subsection{Exploratory analysis}
% TODO: summary stats with sample size
%\subsubsection{Summary statistics}

\subsubsection{Price history}
Figure \ref{fig: price_date} plots the price history of all BAYC trades since April 2021. Figure \ref{fig: price_rarity} plots the price history of BAYC trades by rarity rank. The plots show that the price of BAYC trades has increased significantly since the collection was launched. The price of BAYC trades is positively correlated with rarity rank, with the rarest NFTs trading at a premium.
\begin{figure}[H]
    \caption{Trade Price by Date}
    \includegraphics[width=\textwidth]{../figures/price_date.png}
    \label{fig: price_date}
\end{figure}

\begin{figure}[H]
    \caption{Trade Price by Rarity Rank}
    \includegraphics[width=\textwidth]{../figures/price_rarity.png}
    \label{fig: price_rarity}
\end{figure}

\subsubsection{Principal component analysis}
Figure \ref{fig: pca} plots the first three principal components of the traits and rarity features. The plot shows that the traits and rarity features are not linearly separable, which motivates the use of non-linear models such as tree-based models and deep learning models.   
\begin{figure}[H]
    \caption{PCA(3) plot of NFT prices (Bored Ape Yacht Club)}
    \includegraphics[width=\textwidth]{../figures/pca_1.png}
    \label{fig: pca}
\end{figure}

\section{Methodology}
\label{sec: method}
\subsection{Feature selection and benchmark models}
To find the set of features with the highest predictive power, I define three feature sets and examine the predictive outcome use linear and tree-based models.

Approximately 20\% of the trades dataset feature first-time trades, which lack the last\_trade features. This implies a skewed distribution of trading frequency among BAYCs. Therefore one major goal of this step is to examine the choice of including the last\_trade variables at the cost of dropping that portion of training data.

I define three feature sets, each containing a combination of feature categories:
\begin{itemize}
    \item Set 1: market, traits \& rarity, last trade
    \item Set 2: market, traits \& rarity
    \item Set 3: traits \& rarity, last trade
\end{itemize}

For each feature set, I fit three types of linear models (OLS, Lasso, Ridge) and two types of tree-based ensemble models (random forest, gradient boosting), selecting the feature set that yields the best outcome metric.

As a robustness check, I use the same model-feature set pairs to predict an alternative outcome variable: the log-transformed trade price of each trade.

\subsection{Deep learning models}
I define a class of convolutional neural network (CNN) and use hyperparater tuning to find the architecture that results in minimum loss. The choice for CNN is motivated by the cross-sectional nature of the data. Convolutional layers are designed to capture high-order interations between the traits features and market movements. The CNN model is tuned using random search over the hyperparameters described in Section \ref{sec: hyperparam_tuning}.

\section{Results}
\label{sec: results}
\subsection{Benchmarking}
Benchmarking results using linear and tree-based models are reported in Table \ref{tab:benchmark}
\begin{table}[H]
    \centering
    \caption{MSE Loss Metric for Benchmark Models and Explanatory Variable Sets}
    \label{tab:benchmark}

    \begin{subtable}{\linewidth}
    \centering
    \caption{Y = Trade price (ETH)}
    \begin{tabular}{lccc}
    \hline
    Model & X1 & X2 & X3 \\ 
    \hline
    OLS & 353.02 & 319.01 & 762.74 \\
    Lasso & 374.11 & 327.36 & 772.64 \\
    Ridge & 353.02 & 319.01 & 762.74 \\
    Random Forest & 154.97 & 200.96 & 254.85 \\
    Gradient Boosting & 143.20 & 200.50 & 262.19 \\
    \hline
    \end{tabular}
    \end{subtable}%
    \bigskip

    \begin{subtable}{\linewidth}
    \centering
    \caption{Y = log (Trade price (ETH))}
    \begin{tabular}{lccc}
    \hline
    Model & X1 & X2 & X3 \\ 
    \hline
    OLS & 0.97 & 1.63 & 1.52 \\
    Lasso & 2.07 & 2.73 & 2.73 \\
    Ridge & 0.97 & 1.63 & 1.52 \\
    Random Forest & 0.20 & 0.43 & 0.27 \\
    Gradient Boosting & 0.22 & 0.43 & 0.28 \\
    \hline
    \end{tabular}
    \end{subtable}%
    \bigskip

    \begin{subtable}{\linewidth}
    \centering
    \caption{Y = Trade price (USD)}
    \begin{tabular}{lccc}
    \hline
    Model & X1 & X2 & X3 \\ 
    \hline
    OLS & 5.01e+09 & 4.37e+09 & 8.29e+09 \\
    Lasso & 5.01e+09 & 4.37e+09 & 8.29e+09 \\
    Ridge & 5.01e+09 & 4.37e+09 & 8.29e+09 \\
    Random Forest & 1.29e+09 & 1.27e+09 & 2.91e+09 \\
    Gradient Boosting & 2.03e+09 & 2.29e+09 & 3.23e+09 \\
    \hline
    \end{tabular}
    \end{subtable}
\end{table}

\subsection{Deep Learning Models}

\subsubsection{Model architecture}
The deep learning models are trained using the \texttt{tensorflow.keras} package in \texttt{Python 3.9} under an \texttt{WSL-2 Ubuntu} operating system. Figure \ref{fig:cnn_arch} illustrates the model architecture, without the optional dropout layer.

\begin{figure}[H]
    \centering
    \caption{Example CNN Architecture. The actual architecture is tuned using \texttt{keras\_tuner} package.}
    \label{fig:cnn_arch}
    \includegraphics[width=\textwidth]{../figures/nn.png}
\end{figure}

\subsubsection{Hyperparameter tuning}
\label{sec: hyperparam_tuning}

The tunable hyperparameters include: number of filters and kernel size in the convolutional layer, number of units in the first dense layer, a boolean to include the dropout layer, and the learning rate for Adam optimizer. The feature importance and correlation metrics are computed post-hoc using the weights \& biases package.

\begin{table}[H]
    \centering
    \caption{List of Hyperparameters and Importance with Respect to Validation Loss.}
    \begin{tabular}{l|c|c}
    \hline
    \textbf{Config Parameter} & \textbf{Importance} & \textbf{Correlation} \\ \hline
    kernel size             & .427            & -.102  \\ 
    learning rate           & .356            & .055  \\ 
    filters                  & .100            & .008  \\ 
    units                    & .087            & -.091  \\ 
    dropout                  & .030            & -.098  \\ \hline
    \end{tabular}
    \label{tab:cnn_hyperparam}
\end{table}

I tune these hyperparameters using the \texttt{RandomSearch} method in \texttt{keras\_tuner} package.  During training, the number of epochs is capped at 25, with a batch size of 32. The training and validation loss by epoch is shown in Figure \ref{fig:cnn_loss}. The searched hyperparameters and resulting validation loss (MSE) are shown in Figure \ref{fig:cnn_tuning}.

\begin{figure}[H]
    \centering
    \caption{CNN Training and Validation Loss by Epoch. Each line represents a trial with different hyperparameters}
    \label{fig:cnn_loss}

    \begin{subfigure}{\textwidth}
        \caption{Training loss by epoch}
        \includegraphics[width=\textwidth]{../figures/wandb_loss_nn.png}
    \end{subfigure}
    \begin{subfigure}{\textwidth}
        \caption{Validation loss by epoch}
        \includegraphics[width=\textwidth]{../figures/wandb_val_loss_nn.png}
    \end{subfigure}
\end{figure}

\begin{figure}
    \caption{CNN Hyperparameters Searched using \texttt{keras\_tuner.RandomSearch} method}
    \label{fig:cnn_tuning}
    \includegraphics[width=\textwidth]{../figures/wandb_tuning.png}
\end{figure}

\section{Application}
\label{sec: application}
In this section, I propose an application framework for this model using zero-knowledge machine learning (zkML) and discuss its potential use cases in the context of decentralized applications (DApps).

The goal of the framework is to provide a dynamic and trustless valuation model for NFTs while can be easily integrated with onchain NFT marketplaces and decentralized finance (DeFi) applications.

\subsection{Dynamic valuation}
Dynamic valuation empowers the use of NFTs as collateral for DeFi applications, a recent trend in the NFT market. Borrowing against NFTs is currently limited and risky due to the highly volatile market (e.g. ``pump-and-dumps") and the lack of a reliable valuation model for volatile NFT prices. Most existing implementations rely on human input such as user-defined offers, direct negotiations, and Dutch auctions. The incorporation of a reliable valuation model can greatly improve the efficiency of these applications, and allows for dynamic adjustment of borrowing rates and auto-liquidations to protect lenders.

Valuation for an NFT is conducted as follows. For each NFT in a collection (e.g. BAYC \#8817), a collection-specific model is able to output a real-time valuation in ETH based on 1. the last trade price and time since last trade 2. the traits and rarity of the NFT 3. the current market conditions (e.g. volume, price floor, etc.) Due to the dynamic nature of the model, the valuation of each NFT can be updated in real-time as new trades occur and market conditions change.

\subsection{Trustless valuation}
In the traditional banking and loan sector, valuation models are considered top level enterprise secrets and are never published. Hence all users must rely on trust in the enterprise to fairly value their collateralized property, such as a house.

In decentalized applications, however, such level of trust is difficult to establish. This proposed framework solves the issue of trustless valuation by using zero-knowledge machine learning. zkML allows for verifiable proof that a price valuation is indeed calculated by the model, instead of e.g. a manipulated price oracle. Meanwhile, the proof does not reveal the parameters of the model itself. This solves the problem of trustless valuation while preventing exploitation. Recent works by \citet{liu2021zkcnn,zhang2020zero} have demonstrated the feasibility of zkML in ensemble and deep learning models.

In addition, the design choice of using tabular data for all model inputs generates great benefits for fully onchain applications, since storing and transferring image data onchain is costly.

\subsection{Technical implementation}
The technical implementation of this framework consists of three components: a data pipeline, a family of pre-trained prediction models, and a smart contract. The data pipeline is responsible for collecting and processing data from the Ethereum blockchain. The prediction model is responsible for generating the real-time valuation of each NFT. The smart contract is responsible for emitting the valuation of each NFT and providing zero-knowledge proofs onchain.

@@

This research provides insights into valuation of Non-Fungible Tokens (NFTs), particularly in the context of dynamic markets. I provide an innovative approach to the dynamic valuation of NFTs, using deep learning models and a rich cross-sectional dataset from the Ethereum blockchain and OpenSea. Using the Bored Ape Yacht Club NFTs as a demonstration, the model integrates historical trades, market trends, and traits/rarity features to achieve high accuracy in price prediction.

In real-world applications, the incorporation of zero-knowledge machine learning (zkML) can further enhances the model's utility, opening up new possibilities in decentralized finance (DeFi) applications. This work not only contributes to the field of digital asset valuation but also sets a foundation for future research in dynamic and trustless NFT pricing within the evolving landscape of blockchain technologies and decentralized markets.

@@

% \newpage
% \section{Tables and Figures}
% \label{app: tables_figures}
% \nopagebreak
%\processdelayedfloats
